

#Clustering and classification

This week I learned about clustering and classification. How to cluster observations, how to study which factors affect or justify clustering, how many clusters is appropriate etc?


##2. Data input and summary

This week's data comes from an R package called MASS:

```{r}
# access the MASS package
library(MASS)

# load the data
data("Boston")

# explore the dataset
str(Boston)
summary(Boston)
```

The dataset consist of 14 variables and 506 observations. All variables are numerical. One variable ('chas') is a 1/0, presence/absence dummy variable. The variables describe housing values in suburbs of Boston and factors measured at the suburbs which are thought to be related with housing values. Factors include measures of for example crime rate, access to Charles River, nitrogen oxides concentration, average number of rooms per dwelling, distances to five Boston employment centres, accessibility to radial highways, proportion of blacks by town and median value of owner-occupied homes. The full details can be found [here](https://stat.ethz.ch/R-manual/R-devel/library/MASS/html/Boston.html).


##3. Data exploration

Let's explore graphically the distributions and relations of the data:

```{r}
pairs(Boston)
```

This plot is difficult to read. I'll figure out later how to improve the quality of the output. From the summary table I'm however able to explore the variation and distributions of the variables. 

Here are two dotplots of variables 'crim' (per capita crime rate by town) and 'zn' (proportion of residential land zoned for lots over 25,000 sq.ft.) which show that these two variables are not very evenly distributed:

```{r}
dotchart(Boston$crim)
```
```{r}
dotchart(Boston$zn)
```


Some variables seem correlated. Here's a correlation matrix of the variables:

```{r}
library(corrplot)
library(magrittr)

# calculate the correlation matrix and round it
cor_matrix<-cor(Boston) %>%round(digits=2)

# print the correlation matrix
print(cor_matrix)
```

The above matrix is not very readable as it extends into two separate parts. Let's present the correlations in a nicer way.

```{r}
# visualize the correlation matrix
corrplot(cor_matrix, method="circle",type="upper",cl.pos = "b", tl.pos = "d", tl.cex = 0.6)

```

This plot is easier to read. The bigger the circle the more correlated the variables are. Red indicates negative correlation and blue indicated positive correlation.


##4. Standardize dataset

Some of the variables have very high values and wide distributions. We want to scale all variables because later on it may be difficult to sum or average variables that are on different scales. Scaling can be done to all variables in the dataset as they are all numerical.

```{r}
# center and standardize variables
boston_scaled <- scale(Boston)

# summaries of the scaled variables
summary(boston_scaled)

# class of the boston_scaled object
class(boston_scaled)

# change the object to data frame
boston_scaled<-as.data.frame(boston_scaled)
```

Now all the variables have their mean at zero and their distributions are more moderate.


##4. Create a categorical variable

Next I create a categorical variable of the crime rate in the Boston dataset. I use quantiles as the break points. I drop the old crime rate variable from the dataset.

```{r}
# summary of the scaled crime rate
summary(boston_scaled$crim)

# create a quantile vector of crim and print it
bins <- quantile(boston_scaled$crim)
bins

# create a categorical variable 'crime'
crime <- cut(boston_scaled$crim, breaks = bins, include.lowest = TRUE,label=c("low","med_low","med_high","high"))

# look at the table of the new factor crime
table(crime)

# remove original crim from the dataset
boston_scaled <- dplyr::select(boston_scaled, -crim)

# add the new categorical value to scaled data
boston_scaled <- data.frame(boston_scaled, crime)
```

##4. Divide the data into training and testing sets

For later model evaluation purposes I divide the dataset into training and testing datasets, so that 80% of the data belongs to the train set:

```{r}
##dividing the data into training and testing sets

# number of rows in the Boston dataset 
n <- nrow(boston_scaled)

# choose randomly 80% of the rows
ind <- sample(n,  size = n * 0.8)

# create train set
train <- boston_scaled[ind,]

# create test set 
test <- boston_scaled[-ind,]
```

##5. Fit a linear discriminant analysis on the train set. 

Next I want to know which variables might explain the target variable crime rate. I do a linear discriminant analysis with the categorical crime rate as the target variable and all the other variables in the dataset as predictor variables:

```{r}
# linear discriminant analysis
lda.fit <- lda(crime~., data = train)

# print the lda.fit object
lda.fit

# the function for lda biplot arrows
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "red", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}

# target classes as numeric
classes <- as.numeric(train$crime)

# plot the lda results
plot(lda.fit, dimen = 2,col=classes,pch=classes)
lda.arrows(lda.fit, myscale = 1)

```

Variable 'rad' looks like a strong classifying factor. Also 'zn' and 'nox' are dividing the observations.


##6. Predict

Next I want to use the observations in the test set to predict crime classes. I do this because I want to estimate the "goodness" of my model by comparing predictions to observed "real" data. 

For prediction I use the LDA model on the test data. For comparison I tabulate the results with the crime categories from the test set: 

```{r}
# save the correct classes from test data
correct_classes <- test$crime

# remove the crime variable from test data
test <- dplyr::select(test, -crime)

# predict classes with test data
lda.pred <- predict(lda.fit, newdata = test)

# cross tabulate the results
table(correct = correct_classes, predicted = lda.pred$class)

```

I did the random division of train and test data and predicted the above classes twice. First I got a fairly poor result with more than half of the med_high cases predicted incorrectly. On the second round the results look better (results shown here). Some classes are still incorrectly predicted but at least most of the predictions are correct.

##7. K-means clustering

Next I study the boston data without any classifications and try to cluster the data into groups. Maybe the observations form clusters according to the suburbs. I run k-means algorithm on the dataset, investigate what is the optimal number of clusters and run the algorithm again. 

First I reload the Boston dataset and standardize it. Then I calculate the Euklidean distances between the observations and present a summary of the distances:

```{r}
#standardize the data set
boston_scaled2 <- scale(Boston)

# class of the boston_scaled object
class(boston_scaled2)

# change the object to data frame
boston_scaled2<-as.data.frame(boston_scaled2)

# euclidean distance matrix
dist_eu <- dist(boston_scaled2)

# look at the summary of the distances
summary(dist_eu)
```

Next I run the k-means clustering with 3 centers.

```{r}

# k-means clustering
km <-kmeans(boston_scaled2, centers = 3)

# plot the Boston dataset with clusters
pairs(boston_scaled2[9:14], col = km$cluster)
```

I zoomed in to various parts of the plot and found that when looking at the variable 'tax' it is divided into clusters so that at least the black observations belong clearly to their own group.

I also explored the clustering with 5 centers. The grouping seemed even more arbitrary.

Now, I'm not sure about the best number of clusters so I count the total of within cluster sum of squares (WCSS) and see how it behaves when the number of clusters change:

```{r}

library(ggplot2)

# set values
set.seed(123)

# determine the number of clusters
k_max <- 10

# calculate the total within sum of squares
twcss <- sapply(1:k_max, function(k){kmeans(boston_scaled2, k)$tot.withinss})

# visualize the results
qplot(x = 1:k_max, y = twcss, geom = 'line')
```

The total WCSS drops dramatically at around the value 2. That is the optimal number of clusters for this dataset.

I run the clustering again with 2 centers:

```{r}

# k-means clustering
km <-kmeans(boston_scaled2, centers = 2)

# plot the Boston dataset with clusters
pairs(boston_scaled2[1:6], col = km$cluster)

```

Now the clustering seems better, at least for some variable pairs. But on my opinion, having only two groups doesn't tell much. Maybe it suggests that the residents in Boston are divided into two groups, the wealthy and the poor?

##Bonus: 

Next I perform the LDA again to the boston dataset, this time with clusters (3) as the target variable. By visualizing the results with a biplot I can interpret which variables influence the clustering.

```{r}
boston_scaled3<-boston_scaled2

# k-means clustering
km <-kmeans(boston_scaled3, centers = 3)

klusteri<-km$cluster
class(klusteri)

boston_scaled3<-cbind(boston_scaled3,klusteri)
summary(boston_scaled3)

# linear discriminant analysis
lda.fit2 <- lda(klusteri~., data = boston_scaled3)

# print the lda.fit object
lda.fit2

# the function for lda biplot arrows
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "red", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}

# target classes as numeric
classes <- as.numeric(boston_scaled3$klusteri)

# plot the lda results
plot(lda.fit2, dimen = 2,col=classes,pch=classes)
lda.arrows(lda.fit2, myscale = 1)


```

From these results I would interpret that the variable 'rad' (index of accessibility to radial highways) is the strongest linear separator in this dataset. Although many other variables follow not far behind. 

##Super-Bonus: 

Next I'll draw some 3D plots of the training data. NOTE! You might have to click around to see the figure.

```{r}
model_predictors <- dplyr::select(train, -crime)
# check the dimensions
dim(model_predictors)
dim(lda.fit$scaling)
# matrix multiplication
matrix_product <- as.matrix(model_predictors) %*% lda.fit$scaling
matrix_product <- as.data.frame(matrix_product)

library(plotly)
plot_ly(x = matrix_product$LD1, y = matrix_product$LD2, z = matrix_product$LD3, type= 'scatter3d', mode='markers')

#Set the color to be the crime classes of the train set. Draw another 3D plot where the #color is defined by the clusters of the k-means. How do the plots differ?

plot_ly(x = matrix_product$LD1, y = matrix_product$LD2, z = matrix_product$LD3, type= 'scatter3d', mode='markers',color=train$crime)  

```

I stop the exercise here because I'm having trouble understanding the instructions. I'm able to draw these two 3D plots and crime seems to be a strong separator in the dataset. The last plot should demonstrate the division by clusters. However, I'm not sure anymore should I do the k-means clustering again to the training data and then change the plotting code or could I do it just by modifying the color argument.  I leave it here.


